Enforced Invariant Confluence

### 2018-02-23 doc/paper_redux/iconfluence.tex
Added some examples.
Started iconfluence definition tweaks.

The notion of state-based objects is taken from~\cite{shapiro2011conflict}.

@@

\begin{definition}
  An \defword{invariant} $I: S \to \set{\text{true}, \text{false}}$ is a
  predicate on states.
\end{definition}

\begin{example}
  $(\nats, 0, +)$ is a distributed object, $t(x) = 2x$ is a transaction, and
  $I(x) = x \geq 0$ is an invariant.
\end{example}

@@

\subsubsection{Process-Based}
The process-based model is most similar to the process model described by
Shapiro et al.\ in~\cite{shapiro2011conflict}. We are given a set
$\seq{p}{1}{n}$ of $n$ processors each of which begins with state $s_0$.
Processors can execute a transaction $t \in T$ to transition from state $s$ to
state $t(s)$, or they can send their state $s$ to another processor with state
$s'$ causing the other processor to transition from state $s'$ to state $s
\join s'$.

Consider any two processors $p_i$ and $p_j$ with states $s_i^0, \ldots, s_i^n$
and $s_j^0, \ldots, s_j^m$ such that every state $s_i^0, \ldots, s_i^n, s_j^0,
\ldots, s_j^m$ satisfies $I$. $T$ is invariant-confluent with respect to $I$,
abbreviated \defword{\Iconfluent{}}, if $s_i^n \join s_j^m$ is guaranteed to
satisfy $I$.

\subsubsection{Graph-Based}
The graph-based approach is most similar to the model used by Bailis et al.\
in~\cite{bailis2014coordination}. We are given a directed acyclic graph in
which vertices can have an arbitrary number of children but at most two
parents. The vertices are labelled with states, and the edges are either
labelled with transactions or are unlabelled if they correspond to a join. The
graph is really just an alternate way of representing an execution in the
process-based model, but with duplicate states collapsed into a single vertex.
As with the process-based model, we say that $T$ is \defword{\Iconfluent{}} if
the join of any two invariant satisfying states with invariant satisfying
ancestors is guaranteed to satisfy the invariant.

\subsubsection{Expression-Based}
The expression-based approach deals with expressions built from $s_0$,
transactions in $T$, and $\join$. That is, expressions $e$ are built from the
grammar
\[
  e ::= s_0 \mid t(e) \mid e_1 \join e_2
\]
where $t$ corresponds to a transaction in $T$. We can evaluate an expression,
denoted $eval(e)$, in the obvious way:
\begin{mathpar}
  eval(s_0) \defeq s_0

@@

@@

\begin{definition}
  $T$ is \defword{\Iconfluent}, if for all expressions $e_1$ and $e_2$,
  $\Irec{e_1}$ and $\Irec{e_2}$ implies $I(e_1 \join e_2)$.
\end{definition}

@@

>>>>

The notion of state-based objects is taken from~\cite{shapiro2011conflict}.
Note that $\join$ is an \emph{arbitrary} function. It does not necessarily have
to satisfy any special properties like associativity or commutativity, though
later we will see that often it does.

@@

\begin{definition}
  An \defword{invariant} $I \subseteq S$ is a subset of states. Notationally,
  we say $I(s)$ to mean $s \in I$ and $\lnot I(s)$ to mean $s \notin I$.
\end{definition}

\begin{example}
  $(\nats, 0, +)$ is a distributed object, $t(x) = 2x$ is a transaction, and
  $\setst{x \in \nats}{x \equiv 0 \mod 2}$ is an invariant. Here, $I(0)$ and
  $I(2)$ but $\lnot I(1)$ and $\lnot I(3)$.
\end{example}

@@

\subsubsection{Process-Based}
The process-based model is most similar to the process model described by
Shapiro et al.\ in~\cite{shapiro2011conflict}. We are given a set
$\seq{p}{1}{n}$ of $n$ processors each of which begins with state $s_0$.  A
processor $p_i$ can perform one of two actions.

First, $p_i$ can execute a transaction $t \in T$ to transition from state $s$
to state $t(s)$, given that $I(t(s))$. If $\lnot I(t(s))$, then $p_i$ will not
execute $t$ (or it will execute $t$ but then abort it; think of it however you
like).

Second, $p_i$ can send its state $s_i$ to another processor $p_j$ with state
$s_j$ causing $p_j$ to transition from state $s_j$ to state $s_i \join s_j$ (or
$s_j \join s_i$; it doesn't really matter). Note that unlike executing a
transaction, $p_j$ must transition from $s_j$ to $s_i \join s_j$ even if $\lnot
I(s_i \join s_j)$.

$T$ is invariant-confluent with respect to $I$, abbreviated \Iconfluent{}, if
every reachable state (including $s_0$) satisfies the invariant. Note that this
definition of \Iconfluence{} is different from the original definition
in~\cite{bailis2014coordination}. Later, we will see that they are equivalent.

\subsubsection{Graph-Based}
The graph-based approach is most similar to the model used by Bailis et al.\
in~\cite{bailis2014coordination}. We are given a directed acyclic graph where
vertices are states, and edges are either labelled with transactions or are
unlabelled if they correspond to a join. The graph is really just an alternate
way of representing an execution in the process-based model, but with duplicate
states collapsed into a single vertex. As with the process-based model, we say
that $T$ is \Iconfluent{} if every vertex in the graph satisfies the invariant.

\subsubsection{Expression-Based}
The expression-based approach formalizes the process-based and graph-based
approach using expressions. Given a state based object $(S, s_0, \join)$ and a
set of transactions $T$, we consider expressions generated by the following
grammar:
\[
  e ::= s_0 \mid t(e) \mid e_1 \join e_2
\]
where $t$ corresponds to a transaction in $T$. We can evaluate an expression
$e$, written $eval(e)$, in the obvious way:
\begin{mathpar}
  eval(s_0) \defeq s_0

@@

\begin{definition}
  Intuitively, we say an expression is \defword{reachable} if it can be reached
  in an execution of the process-based model. Formally, we defined a predicate
  $\reachable{\cdot}$ as follows:

  \begin{mathpar}
    \inferrule{ }{\reachable{s_0}}

    \inferrule{\reachable{e} \\ I(t(e))}{\reachable{t(e})}

    \inferrule{\reachable{e_1} \\ \reachable{e_2}}{\reachable{e_1 \join e_2}}
  \end{mathpar}
\end{definition}

\begin{definition}
  $T$ is \defword{\Iconfluent{}} if $\setst{e}{\reachable{e}} \subseteq I$. In
  other words, $T$ is \Iconfluent{} if all reachable states satisfy the
  invariant.
\end{definition}

As we mentioned earlier, this definition of \Iconfluence{} is different than
the original definition presented in~\cite{bailis2014coordination} (though it
is the same as the definition in~\cite{gotsman2016cause}), but they are
(almost) equivalent. We introduce some definitions and then prove this.

@@

Bailis et al.\ defined \Iconfluence{} to mean that expressions recursively
satisfying $I$ are closed under join. That is, $T$ is \Iconfluent{} if
\[
  \forall e_1, e_2.\; \Irec{e_1} \land \Irec{e_2} \implies I{e_1 \join e_2}
\]
If $\lnot I(s_0)$, then \Iconfluence holds vacuously which is a bit silly, so
Bailis et al.\ probably ought to have also added the condition $I(s_0)$ to
their definition of \Iconfluence{}. Doing so, the two definitions become
equivalent.

\begin{claim}\clmlabel{irec_implies_reachable}
  $\Irec{e} \implies \reachable{e}$
\end{claim}
\begin{elidableproof}
  We perform a structural induction on $e$.
  \begin{itemize}
    \item \textbf{Case $s_0$.}
      Axiomatically, $\reachable{s_0}$.

    \item \textbf{Case $t(e)$.}
      $\Irec{t(e)}$, so $I(t(e))$ and $\Irec{e}$. By the inductive hypothesis,
      $\reachable{e}$, so by the definition of $\reachable{\cdot}$,
      $\reachable{t(e)}$.

    \item \textbf{Case $e_1 \join e_2$.}
      $\Irec{t(e)}$, so $\Irec{e_1}$ and $\Irec{e_2}$. By the inductive
      hypothesis, $\reachable{e_1}$ and $\reachable{e_2}$, so by the definition
      of $\reachable{\cdot}$, $\reachable{e_1 \join e_2}$.
  \end{itemize}
\end{elidableproof}

\begin{claim}
  Consider a state based object $(S, s_0, \join)$, a set of transactions $T$,
  and an invariant $I$. The following two are equivalent:
  \begin{enumerate}[\quad(1)]
    \item
      $\setst{e}{\reachable{e}} \subseteq I$

    \item
      $I(s_0)$ and $\Irec{e_1} \land \Irec{e_2} \implies I{e_1 \join e_2}$
  \end{enumerate}
\end{claim}
\begin{elidableproof}
  First, we show that (1) implies (2). Axiomatically, $\reachable{s_0}$, so by
  (1), $I(s_0)$. Let $e_1$ and $e_2$ be arbitrary expressions such that
  $\Irec{e_1}$ and $\Irec{e_2}$. By \clmref{irec_implies_reachable},
  $\reachable{e_1}$ and $\reachable{e_2}$. Thus, $\reachable{e_1 \join e_2}$,
  so by (1), $I(e_1 \join e_2)$.

  Next, we show that (2) implies (1). We prove by structural induction that for
  all $e$, $\reachable{e} \implies \Irec{e}$. From this, (1) is immediate.
  \begin{itemize}
    \item \textbf{Case $s_0$.}
      $I(s_0)$ by (2), so $\Irec{s_0}$

    \item \textbf{Case $t(e)$.}
      $\reachable{t(e)}$, so $\reachable{e}$ and $I(t(e))$. By the inductive
      hypothesis, $\Irec{e}$, so by the definition of $\Irec{\cdot}$,
      $\Irec{t(e)}$.

    \item \textbf{Case $e_1 \join e_2$.}
      $\reachable{e_1 \join e_2}$, so $\reachable{e_1}$ and $\reachable{e_2}$.
      By the inductive hypothesis, $\Irec{e_1}$ and $\Irec{e_2}$. By $(2)$,
      $I(e_1 \join e_2)$. Thus, $\Irec{e_1 \join e_2}$.
  \end{itemize}
\end{elidableproof}

@@

\todo{Define ``reachable'' states and describe that \Iconfluence{} is
equivalent to saying that reachable states are a subset of invariant satisfying
states.}
\todo{Maybe define Iconfluence to say that the initial state satisfies the
invariant?}
